Large-scale spatiotemporal variation in vital rates of an alpine bird

### 2023-12-20 Manuscript_Nater_etal.qmd
Wrote a first (short?) draft for the introduction

-   There is a growing need/demand for indicators for biodiversity, species abundance and demographic rates (see e.g. EBVs)

-   On one hand, these indicators need to be representative at large scales

-   This for e.g. international reporting towards biodiversity targets

-   On the other hand, indicators with decent spatial resolution are also needed

-   The reason for that is that that is the relevant scale for local-level managements and planning

-   This goes both for development decisions and species management and conservation (which typically happen at a county or even municipality level)

-   Besides organisational scale, there is also spatial variation not only in biodiversity, but also in population dynamics and life history of (target) species

-   This variation needs to be accounted for to develop successful and sustainable (harvest) management and conservation strategies

-   There is clear need for large-scale and spatially explicit indicators, but their development hinges on the availability of data from large-scale and long-term monitoring programmes

-   Several countries (in Europe) have been working towards setting up such programmes and been successful particularly for monitoring of some species groups, i.e. breeding birds and butterflies

-   Unfortunately, most of these monitoring programmes te limited to presence-absence or very simple count observations, making them suitable for the development of indicators of distribution and population trends, but usually not for indicators of abundance, population dynamics, and demographic rates

-   The primary reason here is that large-scale long-term monitoring programmes are very costly to run, in particular when they need to be carried out by specialists

-   Citizen science, i.e. the collection of ecological data by members of the public, has been growing rapidly and providing data at larger spatial and taxonomic scales than previously possible.

-   However, increased coverage and total sampling effort in citizen science monitoring often comes at the cost of data being presence-only and prone to bias, limiting usefulness in particular for the study of population dynamics and demography.

-   The exception are monitoring programmes that make use of a large number of citizen scientists that have been trained to collect data and record metadata in a structured manner and according to a carefully designed protocol (+ 1-2 internationally "widely known" programmes?)

-   Norway has one such monitoring programme that focuses on terrestrial gamebird species: Hønsefugltakseringen (needs an English name)

-   Brief intro of monitoring programme: sampling design (mention age- and sex- assigment), coverage, data handling, OA publishing of data

-   Brief intro of previously developed modelling framework: integrated distance sampling model using age-specific line transect data and auxiliary telemetry data to get estimates of population size and structure, as well as key vital rates (survival, recruitment), over time

-   Combined, national-scale monitoring and this modelling framework have potential for large-scale high-resolution modelling of population dynamics and its demographic and environmental drivers

-   In this study, we extended model of Nielsen & Nater 2023 to run not just on a single area but all areas with publicly available line transect data simultaneously

-   Capitalize on "sharing information" across locations and effectively space-for-time substitution to estimate population size, age-structure, survival, and recruitment across 41 reporting districts and 15 years (2007-2021) for willow ptarmigan

@@

-   Willow ptarmigan distribution, life history, and (historical) abundance

@@

The line transect survey data were collected through a structured participatory monitoring program called "Hønsefuglportalen" (<https://honsefugl.nina.no/Innsyn/en>). In August each year, trained volunteer fieldworkers collect willow ptarmigan observations along predefined line transects. To increase the detection probability, each team of fieldworkers use a pointing dog to locate the birds. Each survey team typically consist of at least two persons (one dog handler and one person responsible for following the transect line) and one dog. Often, more than one dog is used for a survey, but only one dog should be used at a time. The transect lines vary in length, but is typically between x and y km (range: xx, mean: yy). When an observation is made, the exact location of observation is noted, along with perpendicular distance to the transect line, and the age- and sex of the observed birds. An observation typically include 1 - 12 birds (mean = ?), representing one brood (female and or male with young-of-the-year chicks). When the surveys are conducted in August, the chicks of the year are able to fly but can be distinguished from older birds as they are still of smaller body size. Since 2019, most of the data has been collected using a mobile app tailored made for the monitoring program, where the field workers can register and get access to the transect lines allocated to them by the local organizers. Before 2019, data were collected on a dedicated field form, and entered manually in a web portal. After field data has been collected, several steps of quality control takes place before, both by local stakeholders and personnel from NINA. Surveys are carried out on both public and private land. After an initial embargo period, all data from public land are published and made freely available as a sampling-event data set on GBIF (refs). This data set contain both meta data about the transect surveys (survey date, line transect length and location, study area ID and other information) as well as data pertaining to the observations (species, number of birds of different categories (adult males, adult females, juveniles, and birds of unknown category), perpendicular distance to transect line, exact location, and time of observation). For this study, we used only data from public land that is freely available on GBIF; thus the workflow we describe should be fully reproducible by others. Formally, the data is publsihed as three distinct data sets, based on the three main public land owners (Statskog, FeFo and Fjellstyrene, respectively).

The program is not designed as a national monitoring, but rather a collection of local and regional survey programs. All involved survey areas use a common field protocol and data collection model. In addition, the local study designs are reviewed by staff at NINA, and common recommendations for study design is provided. However, both a local scale and in particular at a regional scale, the design is unbalanced due to the fact that participation in voluntary by stakeholders. In general, the effort is higher in South-Eastern and Central Norway, medium in Northern Norway, and in general low in Western and Southern Norway. In total, we used data from Xx different survey areas, across xx counties and xx municipalities.

>>>>

There is growing demand for biodiversity indicators from international unions, national government, local management bodies, and corporate and industry actors. Such indicators ideally represent a wide range of biodiversity's states and functions (e.g. + ref on basic EBV idea), yet the development of suitable indicators for certain qualities, such as species abundance and demography, has been more difficult than for others (+ref). This is at least partially due to challenging requirements regarding spatial scales of useful biodiversity indicators. On one hand, indicators need to be representative at large geographic scales, for example, in the context of countries' reporting towards biodiversity targets (+ ref). On the other hand, indicators also ideally have good spatial resolution as the scales relevant for local-level management and planning are often much smaller (+ ref). This latter requirement is particularly crucial for (infrastructure) development strategies and for species management and conservation, both of which tend to require knowledge on species abundance and population dynamics (i.e. demographic rates) that is relevant for their county- or municipality-level decision making (+ refs). 
Another reason why abundance and population indicators ideally come with good spatial resolution is that there can be substantial amounts of variation in population dynamics and life history of species across space (+ref). This is variation that needs to be accounted for in order to develop successful and sustainable strategies for area use, (harvest) management, and species and biodiversity conservation [@williams2002].

While large-scale, spatially-explicit indicators for species abundance and populations are clearly needed, development and practical implementation are greatly limited due to the reliance of such indicators on the availability of data from large-scale, long-term monitoring programmes (+ref). Consequently, many countries have been working on setting up, maintaining, and improving such monitoring programmes over the last decades. Many now well-established programmes focus on breeding birds and butterflies, and examples include the North American Breeding Bird Survey (https://www.pwrc.usgs.gov/bbs/), UK Butterfly Monitoring Scheme (https://ukbms.org/), Swiss Biodiversity Monitoring (https://www.biodiversitymonitoring.ch/), and many more. <!-- is there any other one we want to name here specifically? -->

There is a natural trade-off between quality and quantity of data that can be collected in any monitoring programme: collecting high quality data in a structured manner is costly and requires trained specialists, thus limiting the amount of data that can be collected, while participatory monitoring, i.e. the collection of ecological data by members of the public (also called citizen science, +ref), allows to greatly reduce costs and extend spatial and taxonomic scales of monitoring at the expense of data quality and risk of bias (+ref). Consequently, many large-scale monitoring programmes are limited to presence-absence or very simple count observations, making them suitable for the development of indicators of distribution and population trends, but usually not for indicators of abundance, population dynamics, and demographic rates (+ ref). 
The exception here are monitoring programmes that succeed in making use of a large number of citizen scientists that have been trained to collect data and record metadata in a structured manner and according to a carefully designed protocol. <!-- can we think of 1-2 internationally well-known examples? --> In Norway there is a monitoring programme for terrestrial game bird species called "Hønsefuglportalen" (= "game bird portal", https://honsefugl.nina.no/Innsyn/en). It is a line transect survey programme carried out annually in > xxxx locations across the country (xxxx > transects) by trained volunteers using pointing dogs. The programme has a well developed protocol for recording bird observations, auxiliary data, and relevant metadata (+ref to protocol?) and established routines for quality control and annual releases of publicly available data via GBIF. As such, it is particularly well suited to become part of a workflow for producing and updating abundance and population indicators on an annual basis.

The line transect data from "Hønsefuglportalen" has been used previously for estimating abundance trends of willow ptarmigan (*Lagopus lagopus*) across Norway [e.g. @bowler2020; @nilsen2020] but large-scale estimation of demographic rates underlying these trends has thus far remained as untapped potential of the dataset. @nilsen2023 recently developed a novel integrated distance sampling model (IDSM) which successfully uses age information in line transect data alongside radio-telemetry data to estimate not just abundance, but also survival and recruitment across years. 
In this study, we adapt the model of @nilsen2023 to run not just on a single site but on all areas with publicly available line transect data from "Hønsefuglportalen" simultaneously. Unlike several previous studies applying integrated models for population dynamics to multiple (sub-) populations separately and comparing results [e.g. @robinson2014; @nater2023], we opt for an approach explicitly integrating across space, thus allowing for sharing of information across locations and -- in effect -- space-for-time substitution [e.g. @horswill2019; @morrison2022]. We then apply the resulting multi-area IDSM to "Hønsefuglportalen" data on willow ptarmigan to estimate population size, age-structure, survival, recruitment, and impacts of small rodent abundance across  41 reporting districts and 15 years (2007-2021) for willow ptarmigan. We further embed the modelling workflow in a reproducible, semi-automated pipeline that will greatly facilitate the repeated calculation of abundance and population indicators at different spatial scales as new data is added every year.

@@

@@

The line transect survey data were collected through a structured participatory monitoring program called "Hønsefuglportalen" (<https://honsefugl.nina.no/Innsyn/en>). In August each year, trained volunteer fieldworkers collect willow ptarmigan observations along predefined line transects. To increase the detection probability, each team of fieldworkers use a pointing dog to locate the birds. Each survey team typically consist of at least two persons (one dog handler and one person responsible for following the transect line) and one dog. Often, more than one dog is used for a survey, but only one dog should be used at a time. The transect lines vary in length, but is typically between x and y km (range: xx, mean: yy). When an observation is made, the exact location of observation is noted, along with perpendicular distance to the transect line, and the age and sex of the observed birds. An observation typically include 1 - 12 birds (mean = 5.6), representing one brood (female and or male with young-of-the-year chicks). When the surveys are conducted in August, the chicks of the year are able to fly but can be distinguished from older birds as they are still of smaller body size. Since 2019, most of the data has been collected using a mobile app tailored made for the monitoring program, where the field workers can register and get access to the transect lines allocated to them by the local organizers. Before 2019, data were collected on a dedicated field form, and entered manually in a web portal. After field data has been collected, several steps of quality control takes place before, both by local stakeholders and personnel from NINA. Surveys are carried out on both public and private land. After an initial embargo period, all data from public land are published and made freely available as a sampling-event data set on GBIF (refs). This data set contain both meta data about the transect surveys (survey date, line transect length and location, study area ID and other information) as well as data pertaining to the observations (species, number of birds of different categories (adult males, adult females, juveniles, and birds of unknown category), perpendicular distance to transect line, exact location, and time of observation). For this study, we used only data from public land that is freely available on GBIF; thus the workflow we describe should be fully reproducible by others. Formally, the data is publsihed as three distinct data sets, based on the three main public land owners (Statskog, FeFo and Fjellstyrene, respectively).

The program is not designed as a national monitoring, but rather a collection of local and regional survey programs. All involved survey areas use a common field protocol and data collection model. In addition, the local study designs are reviewed by staff at NINA, and common recommendations for study design is provided. However, both a local scale and in particular at a regional scale, the design is unbalanced due to the fact that participation in voluntary by stakeholders. In general, the effort is higher in South-Eastern and Central Norway, medium in Northern Norway, and in general low in Western and Southern Norway. In total, we used data from 41 different survey areas, across xx counties and xx municipalities.
